Quality Control Filtering and Normalization of methylation C4 data
===================================================================
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(minfi)
library(lumi)
library(knitr)
library(rGithubClient)

library(synapseClient)
synapseLogin()

METHYLATIONMETA_ID <-'syn3156828'
```
### Find and download all of the idat files
I'm using annotations to get the metadata because the new samples have yet to be added to the metadata tables.
```{r downloadChunk, cache=TRUE}
fileList <- synQuery('select id, name,biologicalSampleName,C4_Cell_Line_ID,Diffname_short,BeadChip,Array_position,Row,Column,UID from file where parentId=="syn2653626" and fileType=="idat"')
colnames(fileList) <- gsub("file\\.", "", colnames(fileList))

# Assumes in this directory
dataDir <- "../../data/"
files <- lapply(fileList$id,
                function (x) synGet(x, downloadLocation=dataDir))
```
### Load the methylation data
```{r readChunk, cache=TRUE, dependson="downloadChunk"}
RGSet <- read.450k.exp(base = dataDir)
```
### Add the metadata
As noted above, this code is a temporary fix until the metadata tables are updated with the proper information.
```{r}
# metaTable <- synTableQuery(sprintf("select * from %s", METHYLATIONMETA_ID))
# metaTbl <- metaTable@values
fileList$prettyName <- gsub("_Red.*", "", fileList$name, perl=TRUE)
fileList$prettyName <- gsub("_Grn.*", "", fileList$prettyName, perl=TRUE)

metaTbl <- fileList
metaTbl$name <- NULL
metaTbl$id <- NULL
metaTbl  <- unique(metaTbl)

CELLMETA_ID <- "syn2767694"
cellTable <- synTableQuery(sprintf("select * from %s", CELLMETA_ID))
cellTbl <- cellTable@values

meta <- merge(metaTbl, cellTbl, by="C4_Cell_Line_ID")
rownames(meta) <- meta$prettyName

tmpMeta <- data.frame(prettyName=as.character(union(colnames(RGSet), rownames(meta))))
allMeta <-  tmpMeta %>% 
  left_join(meta, by="prettyName")
rownames(allMeta) <- allMeta$prettyName

RGSet <- RGSet[, allMeta$prettyName]
pData(RGSet) <- allMeta[allMeta$prettyName, ]
sampleNames(RGSet) <- pData(RGSet)$UID
meta <- pData(RGSet)
```
### Filtering
Filter out the data sets that are not public and remove poor quality cell-lines.  Specicically SC12-040 has a problematic karyotype.
```{r filterChunk, cache=TRUE, dependson="downloadChunk"}
idx <- (pData(RGSet)$public) & (pData(RGSet)$C4_Cell_Line_ID != 'SC12-040')
RGSet <- RGSet[, idx]
```
### Compute methylated signals and observe the sample qualities
```{r plot_qualityChunk, fig.width=7, fig.height=6, cache=TRUE, dependson="filterChunk"}
MSet <- preprocessRaw(RGSet)
#Plot quality for datasets
qc <- getQC(MSet)
plotQC(qc)
```
### Detection level per sample
```{r fig.width=7, fig.height=6}
detP <- detectionP(RGSet)
failed <- detP > 0.01
plot(as.factor(pData(RGSet)$BeadChip), colMeans(failed), main='Mean Failure Rate per Sample', xlab='Array', ylab='mean # of probes with detection p value >0.01')
```
There are `r sum(rowMeans(failed)>0.05)` probes with detection p value >0.01 in >5% of samples.  Also it is clear that array 6264488096 has issues along with other samples.
### Perform quantile normalization
This also filters out these samples with low qc metrics.
```{r normalizeChunk, cache=TRUE, dependson="filterChunk"}
pData(RGSet)$sex <- ifelse(pData(RGSet)$Gender == "male", "M", "F")
gRatioSet.quantile <- preprocessQuantile(RGSet, fixOutliers = TRUE,
                                         removeBadSamples = TRUE,
                                         badSampleCutoff = 10.5,
                                         quantileNormalize = TRUE,
                                         stratified = TRUE,
                                         mergeManifest = FALSE, 
                                         sex = NULL)

beta <- getBeta(gRatioSet.quantile)
```
These samples were removed:
```{r, echo=FALSE, results='asis'}
removed <- setdiff(colnames(RGSet), colnames(gRatioSet.quantile))

printCols <- c("UID", "C4_Cell_Line_ID", "Diffname_short")

kable(pData(RGSet)[removed, printCols])
```
### Review the quality of the normalized data with PCA and clustering
```{r, fig.width=9}
#sub-sample the large matrix
smaller <-beta[sample(nrow(gRatioSet.quantile),10000), ]
#Perform PCA and plot color based on 
plotSampleRelation(smaller, method='mds', cv.Th=0, color=pData(gRatioSet.quantile)$Diffname_short)
plotSampleRelation(smaller, method='cluster', cv.Th=0, cex=.9)
#The second PC corresponds to gender
plotSampleRelation(smaller, method='mds', cv.Th=0, color=as.character(pData(gRatioSet.quantile)$Gender), dimension=c(2,3))
#The third PC is somewhat related to lab of origin
plotSampleRelation(smaller, method='mds', cv.Th=0, color=as.character(pData(gRatioSet.quantile)$Originating_Lab), dimension=c(3,4))
```
### Store the normalized dataset in Synapse
```{r echo=FALSE}
repo <- getRepo("kdaily/pcbc_c4_analysis", ref="branch", refName="methylation")
script <- getPermlink(repo, "code/Rmd/normalize_methylation.Rmd")

beta <- cbind(data.frame(ProbeID=rownames(beta)),
              beta)

write.table(beta, file='C4Meth450.tsv', sep='\t', row.names=FALSE, quote=FALSE)

file <- File('C4Meth450.tsv', parentId='syn2813225', name='Normalized Methylation Matrix')
synSetAnnotations(file) <- list(dataType="methylation", fileType="genomicMatrix")
usedFiles <- c('syn2653626')

act <- Activity(name="Normalize", 
                description="Normalize methylation using minfi and lumi", 
                used=usedFiles, executed=script)
generatedBy(file) <- act

data <- synStore(file)
```

```{r echo=FALSE}
detP <- cbind(data.frame(ProbeID=rownames(detP)),
              detP)

write.table(detP, file='detectionPValues.tsv', sep='\t', row.names=FALSE, quote=FALSE)

detfile <- File('detectionPValues.tsv', parentId='syn2813225', name='Detection p-values')
synSetAnnotations(detfile) <- list(dataType="methylation", fileType="genomicMatrix")
usedFiles <- c('syn2653626')

act <- Activity(name="Detection p-values", 
                description="Determine detection significance per probe per sample.", 
                used='syn2653626', executed=(script))
generatedBy(detfile) <- act

data <- synStore(detfile)

```

